Do Non-Cognitive Skills Affect Employment Outcomes of Youth During the School-to-Work Transition?

### 2024-09-28 02_codes/paper_draft.qmd

The success of school-to-work transition, i.e., if a young person finds a decent job, is determined by various factors such as education, training, as well as supply and demand for qualified workforce on the labor market. Those young people who, in addition to being unemployed, also do not study, refer to the most vulnerable group in the school-to-work transition and are defined as not in education, employment, or training (NEET). In the recent past, two events amplified vulnerability of youth in terms of labor market outcomes: the Great recession of 2009 and COVID-19 outbreak. The evidence points out that the financial crisis substantially challenged the integration of young people on the labor market all over the world [@tanveerchoudhry2012; @kelly2015; @verick2011; @verd2019; @montalvao2017]. Young people generally and youth employment particularly were put at the center of the global sustainable development agenda, summarized in Sustainable Development Goal 8 on decent jobs and economic growth.[^1] The global pandemic happened in 2019 amplified the labor market challenges faced by the global community. It "has heightened the level of uncertainty experienced by many young people worldwide and has complicated the developmental tasks associated with the transition to adulthood, including leaving home, completing education, and obtaining full-time employment" [@allmang2022]. The initial pandemic estimates suggested that young people were hit more than others by job loss, fell in working hours and substantial drops of income [@ilo2020youth]. It was estimated that as the result of the pandemic shock in 2020 globally youth employment rates fell by 8.7% in comparison to a 3.7 % drop for adults [@ilo2021].

[^1]: Namely, SDG8.6, which asks countries to report the share of youth Not in Education, Employment, or Training.

Though initially the problems of employment and labor market were in the primary interest of economists, the analysis of available academic literature illustrates that in the recent body of work, school-to-work transition was primarily in the focus of psyhologists, sociologists, and then economists, which marks the interdisciplinary nature of the topic [@blokker2023]. With different disciplines bringing various aspects of employment that were not conventionally regarded by economists, the recent studies point out that non-cognitive skills refer to one of the critical components in school-to-work transition [@lerman2013; @glewwe2017; @zudina2022; @ripamonti2023; @avanesian2024]. These skills encompass characteristics of personality not directly tied to cognitive abilities, including motivation, self-control, and social skills. Research increasingly suggests that non-cognitive skills play a crucial role in determining success in the labor market, influencing both employment opportunities and earning potential [@ferguson2011; @almlund2011a; @almlund2011].

@@

Current study explores the effect of non-cognitive skills on employment outcomes of youth aged 15-29 during school-to-work transition in Russia. The Russian data highlights that youth has the highest unemployment rates: in 2022, they accounted for 26% for 15-19 year olds, 13.6% for 20-24 year olds, and 5.2% for 25-29 year olds. This is compared to 2.8% of unemployment for individuals in the age of 40 to 44 year olds, the lowest value in the sample [@rosstat2024, p. 122]. Further, out of all higher education graduates between 2020-2022 who find a job, for 24% the obtained employment does not match the received profession. Finally, the available national data evidence that the labor market in the country demonstrates unprecedented spatial segregation: with the 4% as a country's average unemployment in 2022, it varies in the range between 1.6% in Yamalo-Nenets Autonomous Okrug and 30% in Republic of Ingushetia.[^2]

[^2]: The Rosstat data were accessed on 24 September 2024, via the following link: <https://rosstat.gov.ru/storage/mediabank/Trud_3_15-s.xlsx>

@@

In order to estimate the effect of non-cognitive skills on employment outcomes, the study relies on the unbalanced panel, where substantial share of observations has only one record. In order to address the research questions that drive this investigation, mixed-effects regression model is adopted. It's primary advantage refers to the possibility to control endogeneity that occurs due to the unobserved heterogeneity at multiple levels.

First, as the data is longitudinal and XX% of the sample has two records in time, the model adopts random intercept ID to separate between- and within-individual variation in employment and non-cognitive skills. Further, the age range of the study is 15-29 in accordance with the ILO definition of the school-to-work transition, but due to differences in baseline probability due to age, it would be the source of endogeneity at the individual level. Younger people, due to school enrollment, would be associated with low incidence rates of employment. With that regard, the model adopted to estimate the effect of non-cognitive skills on employment will have to isolate the effect of age. For that, the model adopts the random intercept for age, which allows the probability of employment to vary across age.

@@

Needs to be taken into account that existing structures of socio-economic inequality make substantial footprint on school-to-work transition of young people, with youth from the richer households having access to better schools, parents being included in networks , which results in having more default options for youth to find a job. With that regard, the poor, who experience greater risks of lacking access to social lifts to elevate their status in the society, might also face higher risks of unemployment due to the lower human capital they possess. With that regard, for the purpose of labor policy it is essentially important to understand which non-cognitive skills could help economically disadvantaged youth have access to jobs. For that, a second model adopts the random slope terms of non-cognitive skills by the income percentile groups.

@@

1\) Quality of data due to the misingness and attrition makes the data potentially highly prone to sample selection bias.

Based on the provided methodology section, here are potential research limitations and suggestions to address them. I've considered the mentioned data missingness and sample selection bias, as well as other potential issues:

1.  **Missing Data and Sample Selection Bias:**

    -   *Limitation*: The data suffers from missingness, which can lead to sample selection bias.

    -   *Suggestion*:

        -   Implement multiple imputation techniques to handle missing data. This involves creating multiple datasets with imputed values for missing observations and then combining the results from analyzing these datasets. R packages like 'mice' or 'Amelia' can be used for this purpose.

        -   Perform sensitivity analyses to assess the impact of different missing data handling strategies on the results.

        -   Consider using inverse probability of censoring weighting (IPCW) or other techniques to address sample selection bias due to missing data.

2.  **Unbalanced Panel and Attrition:**

    -   *Limitation*: The panel is unbalanced, with a substantial share of observations having only one record, and there might be attrition bias.

    -   *Suggestion*:

        -   Investigate the reasons for attrition and compare the characteristics of individuals who dropped out with those who remained in the panel.

        -   Consider using techniques like inverse probability of attrition weighting (IPAW) or shared frailty models to account for attrition bias.

3.  **Endogeneity:**

    -   *Limitation*: The study acknowledges endogeneity due to unobserved heterogeneity at multiple levels (age, region, etc.).

    -   *Suggestion*:

        -   While mixed-effects models help control for endogeneity, consider using instrumental variable (IV) approaches or difference-in-differences (DID) methods for specific research questions where suitable instruments or comparison groups can be identified.

        -   Collect more detailed information in future studies to account for additional unobserved factors.

4.  **Measurement Error:**

    -   *Limitation*: Self-reported data on non-cognitive skills and employment outcomes might be subject to measurement error and social desirability bias.

    -   *Suggestion*:

        -   Consider using objective measures of non-cognitive skills (e.g., cognitive tests, observational studies) or validated scales in future research.

        -   Collect data from multiple sources (e.g., respondents and their parents or teachers) to triangulate information.

5.  **Longitudinal Analysis:**

    -   *Limitation*: The study uses a longitudinal dataset but doesn't explicitly address time-varying covariates or dynamic effects.

    -   *Suggestion*:

        -   Incorporate time-varying covariates in the mixed-effects models to capture changes in individual characteristics over time.

        -   Consider using growth curve models or other advanced longitudinal techniques to analyze dynamic effects.

6.  **Interactions and Non-linear Effects:**

    -   *Limitation*: The study focuses on linear effects and interactions, but non-cognitive skills might have non-linear effects on employment outcomes.

    -   *Suggestion*:

        -   Include quadratic or polynomial terms for non-cognitive skills in the models to capture non-linear effects.

        -   Consider using machine learning techniques or non-parametric approaches to model complex relationships.

By addressing these limitations, you can strengthen the validity and robustness of your research findings.

>>>>

The success of school-to-work transition, i.e., if a young person finds a decent job, is determined by various factors such as education, training, as well as supply and demand for qualified workforce on the labor market. Those young people who, in addition to being unemployed, also do not study, constitute the most vulnerable group in the school-to-work transition and are defined as not in education, employment, or training (NEET). In the recent past, two events amplified vulnerability of youth in terms of labor market outcomes: the Great recession of 2009 and the COVID-19 outbreak. The evidence points out that the financial crisis substantially challenged the integration of young people on the labor market all over the world [@tanveerchoudhry2012; @kelly2015; @verick2011; @verd2019; @montalvao2017]. Sone time later, young people generally and youth employment specifically were put at the center of the global sustainable development agenda, summarized in Sustainable Development Goal 8 on decent jobs and economic growth (e.g., SDG 8.6: Promote Youth Employment, Education and Training). However, the global pandemic happened in 2019 amplified the labor market challenges faced by the global community. It "has heightened the level of uncertainty experienced by many young people worldwide and has complicated the developmental tasks associated with the transition to adulthood, including leaving home, completing education, and obtaining full-time employment" [@allmang2022]. The initial pandemic estimates suggested that young people were hit more than others by job loss, fell in working hours and faced substantial drops of income [@ilo2020youth]. It was estimated that as the result of the pandemic shock in 2020, global youth employment rates fell by 8.7% in comparison to a 3.7 % drop for adults [@ilo2021].

Though initially the problems of employment and labor market were in the primary interest of economists, the analysis of available academic literature highlights the interdisciplinary nature of the topic, with psychology, sociology, and economics being top 3 academic disciplines to focus on the subject [@blokker2023]. With different disciplines bringing various determinants of employment that were not conventionally regarded by economists, the recent studies point out that non-cognitive skills refer to one of the critical components in the school-to-work transition [@lerman2013; @glewwe2017; @zudina2022; @ripamonti2023; @avanesian2024]. These skills encompass characteristics of personality not directly tied to cognitive abilities, including motivation, self-control, social skills, and emotional regulation. Research increasingly suggests that non-cognitive skills play a crucial role in determining success in the labor market, influencing both employment opportunities and earning potential in both developed [@ferguson2011; @almlund2011a; @almlund2011] and developing economies [@rozhkova2019; @gimpelson2020].

@@

Current study explores the effect of non-cognitive skills on employment outcomes of youth aged 15-29 during school-to-work transition in Russia. The Russian data highlights that youth has the highest unemployment rates: in 2022, they accounted for 26% for 15-19 year olds, 13.6% for 20-24 year olds, and 5.2% for 25-29 year olds. This is compared to 2.8% of unemployment for individuals in the age of 40 to 44 year olds, the lowest value in the sample [@rosstat2024, p. 122]. Further, out of all higher education graduates between 2020-2022 who find a job, for 24% the obtained employment does not match the received profession. Finally, the available national data evidence that the labor market in the country demonstrates unprecedented spatial segregation: with the 4% as a country's average unemployment in 2022, it varies in the range between 1.6% in Yamalo-Nenets Autonomous Okrug and 30% in Republic of Ingushetia.[^1]

[^1]: The Rosstat data were accessed on 24 September 2024, via the following link: <https://rosstat.gov.ru/storage/mediabank/Trud_3_15-s.xlsx>

@@

In order to estimate the effect of non-cognitive skills on employment outcomes, the study relies on the unbalanced panel, where substantial share of observations has only one record. In order to address the research questions that drive this investigation, mixed-effects (multilevel) regression model is adopted [@gelman2006; @mixed-ef2000; @wu2009]. These models are particularly useful in estimating effects on non-independent, clustered or hierarchical observations and repeated measures [@yang2014]. Moreover, it's primary advantage refers to the possibility to control endogeneity that occurs due to the unobserved heterogeneity at multiple levels.

First, as the data is longitudinal and XX% of the sample has two records in time, the model adopts random intercept ID to separate between- and within-individual variation in employment and non-cognitive skills. However, to produce unbiased estimates of non-cognitive skills on employment outcomes, the model needs to acknowledge the unobserved heterogeneity in employment due to time-varying characteristics, such as changes in economy, demographics, policy, etc. Random intercept for the survey year would isolate the effect of these factors.

Further, the age range of the study refers to 15-29 year-olds in accordance with the ILO definition of the school-to-work transition, but with respect to differences in baseline probability due to age, it would be the source of endogeneity at the individual level. Lower incidence rates of employment would be observed amongst younger people enrolled at secondary school or higher education institutions. With that regard, the model adopted to estimate the effect of non-cognitive skills on employment will have to isolate the effect of age. For that, the model adopts the random intercept for age, which allows the probability of employment to vary across age.

@@

Needs to be taken into account that existing structures of socio-economic inequality make substantial footprint on school-to-work transition of young people, with youth from the richer households having access to better schools, parents being included in networks , which results in having more default options for youth to find a job. With that regard, the poor, who experience greater risks of lacking access to social lifts to elevate their status in the society, might also face higher risks of unemployment due to the lower human capital they possess. With that regard, for the purpose of labor policy it is essentially important to understand which non-cognitive skills could help economically disadvantaged youth have access to jobs. For that, a second model adopts the random slope terms of non-cognitive skills by the income percentile groups. This is also done to acknowledge a confounding role of SES in the relationship between non-cognitive skills and employment, as the random slope term account for the varying relationships between non-cognitive skills and employment outcomes across different SES levels. Further, the model automatically estimates a random intercept term for SES by non-cognitive skills as well, which captures the unobserved heterogeneity in the relationship between SES and non-cognitive skills.

@@

The study has a number of limitations, both due to the data, as well as estimation strategy. First, data missingness and non-response regarding the income-related questions and non-cognitive skills module create substantial challenges for the unbiased estimation, as the smaller set of observations inevitably results in the sample selection bias. For participants who took part in both waves of the survey, imputation was used to fill the information available in at least one wave.

Attrition bias is also important as the study is based on the unbalanced panel. Though the the effect of time-variant characteristics beyond the scope of the analysis, non-random attrition of data also poses implications on the estimation. Based on longitudinal design, the analysis does not explicitly address time-varying predictors or dynamic effects, except for controlling for the wave/year of the survey through the random intercept term. The further research in this domain could incorporate into the model a number of other time-varying factors that affect both employment outcomes and non-cognitive characteristics.

Finally, though mixed-effects model estimates the effect of non-cognitive skills on employment accounting for the variation at the different levels (e.g., due to age, region, or SES), it is reached via linear terms. Future research that acknowledges non-linearities in personality could help shedding more light on a complex relationship between non-cognitive skills and employment or even a broader set of socio-economic outcomes of an individual.
